A Reduced Order Model for Porous Media Flows: Coupling Linear Regression with Principal Component Analysis (PCA) using SVD or High-Order Singular Value Decomposition (HOSVD) in Upscaling Calculations for Permeability in Reservoir Simulations

### 2018-12-03 Papers/AppliedMaths2018/Resubmission/Stochastic_Upscaling_UK_AMM_2018.tex
Moved Jeff's version to replace my version

\title{A Reduced Order Model for Porous Media Flows which Couples Linear Regression with Principal Component Analysis (PCA) using SVD or HOSVD: Upscaling the Permeability Field for Multiphase Reservoir Flow Simulation}
\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{K. Christou} \author[UoA]{J.L.M.A. Gomes}
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Research Group \\ School of Engineering, University of Aberdeen, UK}

@@

Representation of heterogeneous properties (\ie permeability in this case) within a large system, is a challenge that cuts across many industries that deal with flows in porous media. This is because it is impossible to obtain the value of permeability at every point in realistic domains. And, even if this was possible, it would be extremely challenging to get the computing resources to make use of all available data values. This challenge gave birth to upscaling, a model order reduction method.

In this paper, a single realization of a permeability field was upscaled by first projecting the permeability field into a principal component analysis space (using SVD for 2D field and HOSVD for 3D fields). Then linear regression was employed within each of the principal component spaces to obtain approximate representations of the initial principal component. The new principal components are then recombined to for a new permeability field with a reduced order. This PCA-linear regression method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits/performance.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
Naturally occuring rock formations are inherently heterogeneous, which means that rock properties such as permeability, pore spaces (\ie porosity) and pore throat vary spatially at different length scales. Heterogeneity of rock formations is relevant in calculations of transport and storage of fluids in porous media (\ie multiphase flow in subsurface rocks). Therefore, the representation of such heterogeneous properties is of great importance to oil and gas (\ie reservoir management), carbon capture and storage (CCS, \ie CO$_2$ transportation and storage in subsurface rocks) and waste management (\ie remediation of contaminated soil).

In essence, rock heterogeneity makes it impossible to obtain reliable and accurate spatial information about geological properties throughout the domain (\ie uncertainty problem). Additionally, due to limitations in computing resources, it is neither efficient nor possible to use exact values of geological properties at all spatial coordinates of the domain to perform fine-grid simulations (\ie scaling problem) \cite{chen_2006,miller_1998,Renard_1997}. This is particularly true for large geological domains which span several kilometres. It is possible to overcome these challenges by a family of techniques called {\it upscaling}. Upscaling techniques solve both uncertainty and scaling problems by replacing discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}.

Upscaling techniques which best preserve statistical properties and flow dynamics behaviour of the high resolution domain give the best representation of heterogeneous properties. There are several upscaling techniques which fall under different categories \cite{Hasting_2001, Renard_1997, Szymkiewicz_2013}. Traditional upscaling techniques are deterministic in nature, however over the last decade, techniques classed as stochastic have received great attention from academic and industrial porous media communities worldwide \cite{Guilleminot_2012, Ravalec-Dupin_2010, Verwoerd_2009}. One of the reasons for this is because stochastic upscaling techniques are robust enough to handle multiphase flow in porous media and they are able to address scaling and uncertainity problems.

\medskip
{\it Upscaling} is a term commonly used in the oil and gas sector and is often referred, throughout the applied mathematics and engineering communities, as a form of model order reduction (or reduced order models, MOR/ROM). Depending on the context in which MOR is used, it can simply be referred as dimensionality reduction. The definition of MOR also depends on the context it is being used, but it suffices to state that MOR captures the essential features of a system, structure or parameter \cite{Schilders2008}. This implies that an original system exists with more detailed information. Figure~\ref{fig:IllustrationMOR} presents an interpretation of MOR concept in which a graphical representation of a rabbit is reconstructed with a reduced number of facets (rhs of this illustration).

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998 which introduced Passive Reduced-order Interconnect Macromodeling Algorithm (PRIMA). Truncated Balanced Realization  is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis (PCA) \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and Proper Orthogonal Decomposition (POD) \cite{Sirovich_1987} are also MOR methods discussed in the references published in 1984 and 1987, respectively. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA. In this work, upscaling is performed through linear regression in the PCA space which is obtained with SVD or HOSVD for 2D and 3D cases, respectively.

\bigskip
This work focuses on reducing the order of permeability distribution in porous media flow simulations. Section~\ref{section:mathematical_model} introduces the mathematical model used to describe the relevant physics for multi-fluid flow through porous media and highlights the relevance of the permeability field. The mathematical model is discretised by a novel high-order accurate control volume finite element method (CVFEM) \cite{Gomes_2017} which couples both the fine resolution and upscaled representation of the permeability field. Pre-processing statistical properties and post-simulation multiphase flow behaviour are investigated to assess the various upscaling techniques used. Four upscaling techniques are investigated in this work and they are named Arithmetric mean, Harmonic mean, probability density function (PDF) and PCA (using SVD for 2D models and HOSVD for 3D model) coupled with linear regression.

In Section~\ref{section:pca_linreg}, fundamental mathematics of SVD and associated properties are presented. Then, the concept of spaces is explained to give the reader the understanding necessary to appreciate the transformations and projections involved in this work. Finally, Section~\ref{section:pca_linreg} describes the reduced order model reduction for the permeability field within the principal component space. Section~\ref{section:model_simulation} is dedicated to model description and simulation. It starts by describing the high resolution (\ie base case) on which the four upscaling techniques are applied to. A brief summary of the pre-processing step for each of the models is also presented. Section~\ref{section:results_discussion} provides the results and further discussion on the results of the simulations, it also gives further interpretation to the result obtained from the SVD upscaling technique. Finally, Conclusions are drawn in Section~\ref{section:conclusion}.

\subsection{Brief Overview of Upscaling}\label{section:overview_upscaling}

Upscaling methods replace discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}. This subject of upscaling heterogenous properties has interested industries (\ie oil and gas, CCS and waste management) as well as the academic community for over $7$ decades.

Those interested in large-scale reservoirs or ground water generally take a two-stage geostatistical approach. In the first stage, information from seismic data, well data and analogous outcrops is used to model the large scale heterogeneities associated with facies. In the second stage, statistical models such as continuous multi-variate Gaussian field is used to model the rock properties of the facies \cite{Ewing_1997}. In addition to seismic and well data used in the previous stage, core data can be used as a sources of information for the statistical properties used to determine the random field in this stage.

Several studies were carried out on the upscaling of permeability fields \cite{Christie_2001SPE10Model, Ewing_1997, Hasting_2001, Indelman_1993, King1996, Vereecken_2007, Wen_1996, Yeo2001}. \citet{Cardwell_1945} investigated arithmetic and weighted averages as upscaling techniques for calculating a single equivalent permeability for oil reservoirs with varying permeabilities. The general conclusion reached from this study was that the equivalent permeability lies between the harmonic and the arithmetic average.

\citet{Yeo2001} discussed the accuracy of renormalization method for upscaling 2D hydraulic conductivity fields for two cases. Both cases employed the $4$-quadrant model (see Section~\ref{subsection:model} for details) set-up. The first case had low permeablitity in all the blocks except the bottom right block which had high permeability, while the second case employed the checkboard style described in Section~\ref{subsection:model}. Different conductivity ratios were also investigated for the two cases. The conclusion drawn was that renormalization worked fairly well for the first case, but very poorly for the second case.

\citet{Renard_1997} provided a comprehensive review of the various techniques for upscaling. They found out that the techniques classed broadly under deterministic, stochastic and heuristic complemented each other rather than antagonising each other. \citet{Renard_1997} then recommended that it is better to select a technique which provides block permeabilities instead of one which provides uniform effective permeability.

\citet{Christie_2001SPE10Model} provided $2$ sets of problems to compare the upscaling and upgridding techniques of different simulators. The paper does not describe in any detail the upscaling techniques used by the different simulators, but it provides a complete dataset for running simulations.

@@

PCA, SVD and POD are closely related. In the petroleum engineering community where these techniques are used for reservoir characterization, they are sometimes used interchangable, and are normally referred to as the Karhunen-Loeve Transformation (KLT) \cite{Jafarpour_2009}.

A lot of the recent research in the field of petroleum engineering field which use these techniques, focus on the problem of history matching \cite{Afra_2013, Insuasty_2017, Jafarpour_2009, Tavakoli_2010, Xiao_2018}. History matching is the process of building or adjusting one or more sets of models for a reservoir, such that the model(s) closely reproduces the past behaviour (flow rate, production volume or pressures) of a reservoir, and hopefully predict its future behaviour as well \cite{Rwechungura_2011}. It is important to note that the model(s) referred to here could be a numerical model representing the system of parameters (\ie{variables}) which characterise the reservoir, or simply a model for parameterizing one of more components (\ie{variables}) of the system.

\citet{Jafarpour_2009} compares KLT to discrete cosine transform (DCT) and goes on to state that the DCT is computationally more efficient than the KLT while being almost as accurate. \citet{Tavakoli_2010} noted that history matching calculation of sensitivities for all production data with respect to system parameters (\ie{variables}) is not feasible. Hence, they (\citet{Tavakoli_2010}) explored a new parameterization for reducing the number of variables based on the principal right singular vectors of the dimensionless sensitivity matix.

Perhaps, the works of \citet{Afra_2013}, \citet{Afra_2016}, \citet{Afra_2014}, \citet{Insuasty_2017} are more closely related to the one performed here. However, similar to the work of \citet{Jafarpour_2009}, they are more related to comparing several realizations than reducing the dimensionality of the dataset for a single realization. The work of \citet{Xiao_2018} is also closely related to that performed here and interestingly, they used the same CVFEM simulator used in this work. However, like the work of \citet{Tavakoli_2010}, the \citeauthor{Xiao_2018} were more interested in reducing the number of variable used to describe the system.

Essentially, the recent developments in the field of petroleum engineering described here make use of SVD to select the most suitable realization from multiple realizations or to reduce the representative system variables. However, the work described here is concerned with reducing dimensionality of the dataset for a single realization of one of the system variable (\ie{permeability}).

@@

Immiscible multi-fluid flows in porous media may be represented by the extended Darcy equation,
\begin{equation}\label{e:darcy_eqn}
  \mathbf{q}_{\alpha} = \mathbf{u}_{\alpha}S_\alpha=
  -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left(
  \nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
\end{equation}
where $\alpha$ designates a phase, $\mathbf{q}_{\alpha}$ is the $\alpha$-th phase Darcy velocity and $\mathbf{u}_{\alpha}$ is the saturation-weighted Darcy velocity. $\mu_{\alpha}$, $p_{\alpha}$, $\rho_{\alpha}$, and $\mathbf{s}_{{u}_\alpha}$ are the phase dynamic viscosity, pressure, density and source term, which may include gravity and/or capillarity, respectively. $\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ is the phase relative permeability, which is a function of the phase saturation $S_{\alpha}\left(\mathbf{r},t\right)$, which in turn is a function of the vector position $\mathbf{r}$, and time, $t$.

In addition to the extended Darcy equation (Eqn.~\ref{e:darcy_eqn}), saturation conservative equations,
\begin{equation}
  \phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} + \nabla
  \cdot \left( {\mathbf u}_{\alpha} S_{\alpha}\right) =
  s_{cty,\alpha},
  \label{saturation_equation}
\end{equation}
are discretised in space with CV basis functions, and in time with the $\theta$-method \cite{pavlidis_2013b}. In Eqn.~\ref{saturation_equation}, $\phi$ is the porosity and $s_{cty}$ is a source term.

\medskip
The discretised saturation and Darcy equations are solved using a multigrid-like approach as described by \citeauthor{pavlidis2016}~\cite{pavlidis2016}. The numerical formulation is fully described by \citeauthor{Gomes_2017}~\cite{Gomes_2017} (see also \cite{adam_2016, salinas2015}). These numerical methods are embedded in the next-generation flow simulator Fluidity/IC-FERST model software\footnote{\href{http://multifluids.github.io}{http://multifluids.github.io}} (a full description of the model can be found in \cite{Gomes_2017, jackson_2013}). Although, the numerical formulation used here is relatively new, it is not the focus of this work. The result and discussion section (Section \ref{section:results_discussion}) refers to \citet{dawe_2008} for validation of the numerical formulation with respect to the 2D simulation. However, more extensive validation tests can be found in the aforementioned references \cite{adam_2016, Gomes_2017,jackson_2013, pavlidis2016, salinas2015}.

\medskip
Finally, $\mathbf{K}\left(\mathbf{r}\right)$, the focus of this work, is the absolute permeability tensor of the porous medium and it is a function of the vector position $\mathbf{r}$. $\mathbf{K}$ is prescribed in the system of governing conservative equations, \ie it is an indepedent input variable. Since $\mathbf{K}$ is a petrophysical rock property that varies spatially, it cannot be determined at every spatial coordinate. And, even if this information is available at every spatial coordinate, it would require immense computer resources to process during flow simulations. This is the reason upscaling is required for processing permeability variables. Upscaling can be performed from the continuum scale, across the micro scale ($mm$), local scale and meso scale ($m$) to the field scale ($km$)\cite{ECMI_2004}. The upscaling discussed here is from the micro scale also known as the Darcy scale.

\section{PCA coupled with Linear Regression}\label{section:pca_linreg}

@@

\subsubsection{Understanding the permeability space}\label{subsubsection:visualization_permspace}
Permeability can be represented as tensor quantity, but for the purpose of the following explanation (and the remaining of this paper), it is assumed that permeability is a scalar quantity. Furthermore, the simulator prescribes the permeability field by assigning permeability values to the centre of element-pairs \cite{Christou_2018} which discretises the simulation domain.

Permeability values assigned in the centre of the element/cell represents values of permeability for the associated element $\left(\text{\ie P}_{0}\text{DG element-pair}\right)$. For simplicity, in the remaining part of this sub-section, a $2D$ geometry is assumed where the geometry is discritized by a square grid with the value for each square element assigned in the centre of the element. This explanation allow one to easily visualize this permeability space because it is defined by the cartesian coordinates system.

@@

\subsubsection{Understanding the principal component spaces}\label{subsubsection:visualization_pcspaces}
Essentially, the existing permeability space is transformed with SVD for 2D permeability field and HOSVD for 3D permeability fields, to principal component spaces (PCS). Within the PCS, traditional PCA dimensionality reduction can be optionally performed (\ie{feature selection or elimination}, see Section~\ref{subsubsection:pca_dimred_linreg} for more details), before interpolation is used to implement further data reduction in the principal component spaces. The reduced data in the PCS is then transformed back into the permeability space to obtain a reduced permeability space. An algorithm which impliments the procedure described here, is presented in Section~\ref{subsubsection:svdcase_preprocess_algorithm}

The PCS is a bit challenging to visualize because it represents projections which are embedded in the SVD/HOSVD factors of the original permeability field. An aid for visualizing the PCS for a 2D permeability field is to write out the $\mathbf{U}$, $\mathbf{\Sigma}$ and $\mathbf{V^{\intercal}}$ matrices (or SVD factors) and consider the singular vectors (\ie columns in $\mathbf{U}$ and rows in $\mathbf{V^{\intercal}}$) as separate PCS. A similar intuition can be extended to HOSVD for 3D permeability fields. The decomposition of a matrix into its factors using SVD  \cite{Cline_2007, Tharwat_2016} or HOSVD \cite{Kolda_2009, Lathauwer_2000} is covered in details in several published materials. However, for completeness, the appendix contains a brief introduction to SVD (\ref{subsection:svd_brief}) and HOSVD (\ref{subsection:hosvd_brief}).

\subsubsection{Reduced Order Model/Principal Component Space Reduction}\label{subsubsection:pca_dimred_linreg}
For the PCA-Linear regression (PCA-LinReg) reduced order model of the permeability field, all the work is done in the principal component spaces. This can be completed in two steps, which is exemplified for the 2D model thus:
\begin{enumerate}
  \item The first step is the principal component analysis (PCA) \cite{Hotelling_1933, Tharwat_2016} using the SVD method, where the singular values are examined and, based on a pre-defined criteria, $z$ singular values are selected, starting from $\sigma_{1}$ and in a decreasing order, where $z \leq r$.

>>>>

%\title{A Reduced Order Model for Porous Media Flows which Couples Linear Regression with Principal Component Analysis (PCA) using SVD or HOSVD: Upscaling the Permeability Field for Multiphase Reservoir Flow Simulation}
\title{A Reduced Order Model for Porous Media Flows: Coupling Linear Regression with Principal Component Analysis (PCA) using SVD or High-Order Singular Value Decomposition (HOSVD) in Upscaling Calculations for Permeability in Reservoir Simulations}
\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{K. Christou} \author[UoA]{J.L.M.A. Gomes}
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Research Group \\ School of Engineering, University of Aberdeen, UK}

@@

Representation of heterogeneous petro-physical properties (\eg permeability) within a large system, is a challenge that cuts across many industries that deal with flows in porous media. This is because it is impossible to obtain the value of permeability at every point in realistic domains. And, even if this was possible, it would be extremely challenging to get the computing resources to make use of all available data values. This challenge gave birth to upscaling, a model order reduction method.

In this paper, a single realization of a permeability dataset was upscaled by first projecting the permeability field into a principal component analysis space (using SVD and HOSVD for 2- and 3-D geometries, respectively). Then linear regression was employed within each of the principal component spaces to obtain approximate representations of the initial principal component. The new principal components are then recombined to form a new permeability dataset with a reduced order model. This PCA-linear regression method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits/performance.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
Naturally occuring rock formations are inherently heterogeneous, which means that geophysical properties, such as permeability, pore spaces (\ie porosity) and pore throat vary spatially at different length scales. Heterogeneity of rock formations is relevant in calculations of transport and storage of fluids in porous media (\ie multiphase flow in subsurface rocks). Therefore, the representation of such heterogeneous properties is of great importance to oil and gas (\ie reservoir management), carbon capture and storage (CCS, \ie CO$_2$ transportation and storage in subsurface rocks) and waste management (\ie remediation of contaminated soil).

In essence, rock heterogeneity makes it impossible to obtain reliable and accurate spatial information about geological properties throughout the domain (\ie uncertainty problem). Additionally, due to limitations in computing resources, it is neither efficient nor possible to use exact values of geological properties at all spatial coordinates of the domain to perform fine-grid simulations (\ie scaling problem) \cite{chen_2006,miller_1998,Renard_1997}. This is particularly true for large geological domains which span several kilometres. It is possible to overcome these challenges by a family of techniques called {\it upscaling}. Upscaling techniques solve both uncertainty and scaling problems by replacing discrete geological and fluid properties of detailed high-resolution domains with coarse descriptions (\ie low-resolution) of these properties \cite{Vereecken_2007}.

Upscaling techniques which best preserve statistical geophysical and thermo-physical properties of high-resolution domains result in optimal prediction of fluid flow dynamics. Over the past 60 years, several upscaling techniques were developed (see \cite{Hasting_2001, Renard_1997, Szymkiewicz_2013}). Traditional upscaling techniques are deterministic in nature, however over the last decade, techniques classed as stochastic have received great attention from academic and industrial porous media communities worldwide \cite{Guilleminot_2012, Ravalec-Dupin_2010, Verwoerd_2009}. One of the reasons for this is because stochastic upscaling techniques are robust enough to handle multiphase flow in porous media and they are able to address scaling and uncertainity problems.statistical

\medskip
{\it Upscaling} is a term commonly used in the oil and gas sector and is often referred, throughout applied mathematics and engineering communities, as a form of model order reduction (or reduced order models, MOR/ROM). Depending on the context in which MOR is used, it can be simply referred as `dimensionality reduction'. The definition of MOR also depends on the context in which it is used, but it suffices to state that MOR's are able to capture the essential features of a system, structure or parameter \cite{Schilders2008}. This implies that an original system exists with more detailed information. Figure~\ref{fig:IllustrationMOR} presents an interpretation of MOR concept in which a graphical representation of a rabbit is reconstructed with a reduced number of facets (rhs of this illustration).

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998, which introduced the passive reduced-order interconnect macromodeling algorithm (PRIMA). Truncated balanced realization is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis (PCA) \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and proper orthogonal decompositions (POD) \cite{Sirovich_1987} are also MOR methods discussed in the references published in 1984 and 1987, respectively. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA.

\bigskip
 In this work, upscaling is performed through linear regression in the PCA space which is obtained with high-order singular value decomposition (HOSVD/SVD) in 2-/3-D problems. This work focuses on reducing the order of permeability distribution for porous media flow simulations.

\bigskip
A brief overview of current upscaling methods in reservoir simulation is described in Section~\ref{section:general_upscaling}. Section~\ref{section:mathematical_model} introduces the mathematical model used to describe the relevant physics for multi-fluid flow through porous media and highlights the relevance of the permeability field. The mathematical model is discretised by a high-order accurate control volume finite element method (CVFEM) \cite{Gomes_2017}. Pre-processing statistical properties and post-simulation multiphase flow behaviour are investigated to assess the various upscaling techniques used. Four upscaling techniques are investigated in this work and they are named arithmetric mean, harmonic mean, probability density function (PDF) and PCA (using SVD and HOSVD methods) coupled with linear regression. In Section~\ref{section:pca_linreg}, fundamental mathematics of SVD and associated properties are presented. Then, the concept of spaces is explained to give the reader the understanding necessary to appreciate the transformations and projections involved in this work. Finally, Section~\ref{section:pca_linreg} describes the reduced order model reduction for the permeability field within the principal component space. Section~\ref{section:model_simulation} is dedicated to model description and simulation. It starts by describing the high-resolution problem (\ie `base case') in which the four upscaling techniques are applied to. A brief summary of the pre-processing steps for each of the models is also presented. Section~\ref{section:results_discussion} provides the results and further discussion of the results, it also gives further interpretation to the result obtained from the SVD upscaling technique. Finally, conclusions are drawn in Section~\ref{section:conclusion}.

\section{Upscaling}\label{section:general_upscaling}
\subsection{Brief Overview of Upscaling}\label{section:overview_upscaling}

Upscaling methods replace discrete geological and fluid properties of detailed (high-resolution) domains with coarse descriptions (low-resolution) of these properties \cite{Vereecken_2007}. Upscaling of heterogeneous properties has attracted the interest of several porous media flow communities (\eg oil and gas, CCS and waste management) worldwide for over $7$ decades. Those interested in large-scale reservoirs or groundwater generally take a two-stage geo-statistical approach.

In the first stage, information from seismic data, well data and analogous outcrops is used to model the large-scale heterogeneities associated with facies. In the second stage, statistical models such as continuous multi-variate Gaussian field is used to model rock properties of the facies \cite{Ewing_1997}. In addition to seismic and well data used in the previous stage, core data can be used as a source of information for the statistical properties used to determine random fields in this stage.

A number of studies were carried out on upscaling of subsurface permeability (see \cite{Christie_2001SPE10Model, Ewing_1997, Hasting_2001, Indelman_1993, King1996, Vereecken_2007, Wen_1996, Yeo2001} for a few key references). \citet{Cardwell_1945} investigated arithmetic and weighted averages as upscaling techniques for calculating a single equivalent permeability for oil reservoirs with varying permeabilities. The general conclusion reached from this study was that the equivalent permeability lies between harmonic and arithmetic averages.

\citet{Yeo2001} discussed the accuracy of renormalization method for upscaling 2D hydraulic conductivity fields for two cases. Both cases employed the $4$-quadrant model (see Section~\ref{subsection:model} for details) set-up. The first case had low permeability in all blocks except the bottom right block which had high permeability, while the second case employed the check-board style described in Section~\ref{subsection:model}. Different conductivity ratios were also investigated for the two cases. The conclusion drawn was that renormalization worked fairly well for the first case, but very poorly for the second case.

\citet{Renard_1997} provided a comprehensive review of the various techniques for upscaling. They found out that the techniques classed broadly under deterministic, stochastic and heuristic complemented each other rather than antagonising each other. They also recommended to select a technique which provides block permeabilities instead of one which provides uniform effective permeability. Finally, \citet{Christie_2001SPE10Model} provided $2$ sets of problems to compare the upscaling and upgridding (\ie mesh coarsening) techniques of different simulators. The paper does not describe in any detail the upscaling techniques used by the different simulators, but it provides a complete dataset for flow simulation models' benchmarking.

@@

PCA, SVD and POD are closely related. In the petroleum engineering community where these techniques are used for reservoir characterization, they are sometimes used interchangable, and are normally referred to as Karhunen-Loeve Transformations (KLT) \cite{Jafarpour_2009}.

A number of recent research in reservoir engineering focus on  history matching (HM) problems (see \cite{Afra_2013, Insuasty_2017, Jafarpour_2009, Tavakoli_2010, Xiao_2018}). History matching is the process of building or adjusting one or more sets of reservoir configurations (or realisations), such that reservoir models are able to closely reproduces past flow behaviour (\eg flow rate, production volume, pressure etc) in a geological formation, aiming to predict future flow behaviour with good accuracy \cite{Rwechungura_2011}. It is important to note that reservoir configurations could be either a numerical model representing the system of parameters (\ie{variables}) which characterise the reservoir, or simply a model for parameterizing one of more components (\ie{variables}) of the system.

\citet{Jafarpour_2009} compares KLT and discrete cosine transform (DCT) and concluded that the DCT is computationally more efficient than the KLT while being almost as accurate. \citet{Tavakoli_2010} noted that HM calculation of sensitivities for all production data with respect to system parameters is not feasible. Hence, they explored a new parameterization for reducing the number of variables based on the principal right singular vectors of the dimensionless sensitivity matrix.

Perhaps, work performed by \citet{Afra_2013,Afra_2016,Afra_2014} and \citet{Insuasty_2017} are more closely related to the one performed here. However, similar to the work of \citet{Jafarpour_2009}, they are more related to comparing several realizations than reducing the dimensionality of the dataset for a single realization. The work of \citet{Xiao_2018} is also closely related to that performed here and interestingly, they used the same flow simulator model used in this work. However, similarly to \citet{Tavakoli_2010}, \citeauthor{Xiao_2018} focused on reducing the number of variable used to describe the system.

Essentially, recent developments in the field of reservoir simulation described in this section make use of SVD theory to select the most suitable realization from multiple realizations or to reduce the representative system variables. However, the work described here is concerned with reducing dimensionality of the dataset for a single realization of a single system variable (\ie{permeability}).

@@

Immiscible multi-fluid flows in porous media may be represented by the extended Darcy equation,
\begin{equation}\label{e:darcy_eqn}
  \mathbf{q}_{\alpha} = \mathbf{u}_{\alpha}S_\alpha=
  -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left(
  \nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
\end{equation}
where $\alpha$ designates a phase, $\mathbf{q}_{\alpha}$ is the $\alpha$-th phase Darcy velocity and $\mathbf{u}_{\alpha}$ is the saturation-weighted Darcy velocity. $\mu_{\alpha}$, $p_{\alpha}$, $\rho_{\alpha}$, and $\mathbf{s}_{{u}_\alpha}$ are phase dynamic viscosity, pressure, density and source term (which may include gravity and/or capillarity), respectively. $\mathbf{K}\left(\mathbf{r}\right)$ and $\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ are absolute and phase-relative permeability, respectively. The later is a function of the phase saturation $S_{\alpha}\left(\mathbf{r},t\right)$, which in turn is spatial- and time-dependent.

In addition to the extended Darcy equation (Eqn.~\ref{e:darcy_eqn}), saturation conservative equations,
\begin{equation}
  \phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} + \nabla
  \cdot \left( {\mathbf u}_{\alpha} S_{\alpha}\right) =
  s_{cty,\alpha},
  \label{saturation_equation}
\end{equation}
are discretised in space with CV basis functions, and in time with the $\theta$-method \cite{gomes_book_2012}. In Eqn.~\ref{saturation_equation}, $\phi$ is the porosity and $s_{cty}$ is a source term.

\medskip
The discretised saturation and Darcy equations are solved using a multigrid-like approach as described by \citeauthor{pavlidis2016}~\cite{pavlidis2016}. The numerical formulation is fully described by \citeauthor{Gomes_2017}~\cite{Gomes_2017} (see also \cite{adam_2016, salinas2015}). These numerical methods are embedded in the next-generation flow simulator Fluidity/IC-FERST model software\footnote{\href{http://multifluids.github.io}{http://multifluids.github.io}} (a full description of the model can be found in \cite{Gomes_2017, jackson_2013,salinas_2018}). Although, the numerical formulation used here is relatively new, it is not the focus of this work. The result and discussion section (Section \ref{section:results_discussion}) refers to \citet{dawe_2008} for validation of the numerical formulation with respect to the 2D simulation. However, extensive model and software quality assurance (\ie verification and validation tests) can be found in the aforementioned references \cite{adam_2016, Gomes_2017,jackson_2013, pavlidis2016, salinas2015}.

\medskip
Finally, $\mathbf{K}\left(\mathbf{r}\right)$, the focus of this work, is prescribed in the system of governing conservative equations, \ie it is an independent input variable. Since $\mathbf{K}$ is a petrophysical property of geological formations that varies spatially, it cannot be determined at every grid point (\ie spatial coordinate). And, even if this information is available at every spatial coordinate, it would require immense computer resources to process during flow simulations. This is the reason upscaling is required for processing permeability spatial variability. Upscaling can be performed from the continuum scale, across the micro-scale ($mm$), local- and meso-scale ($m$) to field-scale ($km$)\cite{ECMI_2004}. The upscaling discussed here is for micro-scale, also known as the Darcy scale.

\section{Coupled PCA and Linear Regression}\label{section:pca_linreg}

@@

\subsubsection{Understanding the permeability space}\label{subsubsection:visualization_permspace}
Permeability can be represented as either tensor or scalar quantity, however for the purpose of the following explanation (and the remaining of this work), permeability is assumed as a scalar quantity. In addition, in the numerical simulations performed for this work, permeability is discretised element-wise with projection to the control volume space within each element-pairs (see \cite{Christou_2018}). Permeability values are assigned to the centre of the FE $\left(\text{\ie P}_{0}\text{DG element-pair}\right)$. For simplicity, in the remaining part of this sub-section, a $2D$ geometry is assumed which is discretized with triangular elements (\PN[1]{2} and \PNDG[1]{1} element-pairs). This explanation allows one to easily visualize this permeability space because it is defined by Cartesian coordinate system.

@@

\subsubsection{Understanding the principal component spaces}\label{subsubsection:visualization_pcspaces}
Essentially, the existing permeability space is transformed (with SVD and HOSVD for 2- and 3-D permeability data, respectively) to principal component spaces (PCS). Within the PCS, traditional PCA dimensionality reduction can be optionally performed (\ie{feature selection or elimination}, Section~\ref{subsubsection:pca_dimred_linreg}), before interpolation is used to implement further data reduction in PCS. The reduced data in the PCS is then transformed back into the permeability space to obtain a reduced permeability space. An algorithm which impliments the procedure described here is presented in Section~\ref{subsubsection:svdcase_preprocess_algorithm}

The PCS is a particularly challenging to visualize as it represents projections which are embedded in SVD/HOSVD factors of the original permeability dataset. An aid for visualizing the PCS for a 2D permeability data is to write out $\mathbf{U}$, $\mathbf{\Sigma}$ and $\mathbf{V^{\intercal}}$ matrices (or SVD factors) and consider the singular vectors (\ie columns in $\mathbf{U}$ and rows in $\mathbf{V^{\intercal}}$) as separate PCS. A similar procedure can be extended to HOSVD for 3D permeability dataset. The decomposition of a matrix into its factors using SVD \cite{Cline_2007, Tharwat_2016} or HOSVD \cite{Kolda_2009, Lathauwer_2000} is covered in details in several published materials. However, for completeness, \ref{subsection:svd_brief} and \ref{subsection:hosvd_brief} contain brief overviews of SVD and HOSVD, respectively.
